---
title: "From Model to Report: How tidylearn Simplifies ML Reporting"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## The Reporting Problem

Machine learning in R is powerful, but reporting the results often takes more
effort than building the model itself. Every package returns results in a
different format — a named vector here, a matrix there, a custom S3 object
somewhere else. Even with excellent tidying packages like `broom`, the column
names and available fields change between model types, and many ML packages
aren't supported at all.

The result is that producing consistent, polished visualisations and tables for
a report means writing custom extraction and reshaping code for each package you
use. Change your model, and your reporting code breaks.

**tidylearn** solves this. Every function — across 20+ algorithms — returns tidy
tibbles, ggplot2 plots, and formatted `gt` tables with a consistent structure.
Your reporting pipeline becomes model-agnostic: swap the algorithm, and the same
plot code, table code, and comparison logic work without modification.

This post walks through real-world analysis tasks — PCA, hierarchical
clustering, regularisation, and multi-model classification — comparing the
tidylearn workflow against the traditional approach.

```{r setup}
library(tidylearn)
library(dplyr)
library(ggplot2)
library(gt)
library(tibble)
```

---

## 1. PCA: Biplot, Scree Plot, and Variance Tables

PCA is a staple of exploratory analysis, and producing a polished biplot or
scree plot is one of those tasks that should be straightforward but rarely is.

### With tidylearn

```{r tl-pca}
pca <- tidy_pca(USArrests, scale = TRUE)
```

Scree plot with cumulative variance line and 80% threshold:

```{r tl-pca-scree}
tidy_pca_screeplot(pca)
```

Publication-ready biplot with observation scores and variable loadings:

```{r tl-pca-biplot}
tidy_pca_biplot(pca, label_obs = TRUE)
```

And the tables — variance explained and loadings — are one call each, with
colour-coded formatting out of the box:

```{r tl-pca-variance-table}
pca_model <- tl_model(USArrests, method = "pca")
tl_table_variance(pca_model)
```

```{r tl-pca-loadings-table}
tl_table_loadings(pca_model, n_components = 2)
```

The loadings table uses a diverging red–blue colour scale to highlight strong
positive and negative loadings — no manual formatting required.

### Without tidylearn

```{r base-pca}
pca_base <- prcomp(USArrests, scale. = TRUE)
```

The default R biplot:

```{r base-pca-biplot, fig.height = 6}
biplot(pca_base)
```

This produces a functional but visually rough base R graphic — no `theme_minimal()`,
no consistent colour scheme, no variance-explained axis labels. To get a
ggplot2 biplot, you need to manually extract and scale the scores and loadings:

```{r base-pca-ggbiplot}
# Extract scores
scores <- as.data.frame(pca_base$x[, 1:2])
scores$label <- rownames(scores)

# Extract loadings and scale to match scores
loadings <- as.data.frame(pca_base$rotation[, 1:2])
loadings$variable <- rownames(loadings)
score_range <- max(abs(scores[, 1:2]))
loading_range <- max(abs(loadings[, 1:2]))
scale_factor <- (score_range / loading_range) * 0.8

loadings$PC1_scaled <- loadings$PC1 * scale_factor
loadings$PC2_scaled <- loadings$PC2 * scale_factor

# Compute variance explained for axis labels
var_exp <- pca_base$sdev^2 / sum(pca_base$sdev^2) * 100

# Build the ggplot manually
ggplot() +
  geom_point(data = scores, aes(x = PC1, y = PC2),
             colour = "steelblue", alpha = 0.7) +
  geom_text(data = scores, aes(x = PC1, y = PC2, label = label),
            size = 2.5, vjust = -0.5) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1_scaled, yend = PC2_scaled),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "red", alpha = 0.7) +
  geom_text(data = loadings,
            aes(x = PC1_scaled, y = PC2_scaled, label = variable),
            colour = "red", size = 3, fontface = "bold", vjust = -0.5) +
  labs(
    title = "PCA Biplot",
    x = sprintf("PC1 (%.1f%% variance)", var_exp[1]),
    y = sprintf("PC2 (%.1f%% variance)", var_exp[2])
  ) +
  coord_equal() +
  theme_minimal()
```

```{r base-pca-table}
# Variance table — manually computed
var_explained <- data.frame(
  component = paste0("PC", seq_along(pca_base$sdev)),
  sdev = pca_base$sdev,
  variance = pca_base$sdev^2,
  prop_variance = pca_base$sdev^2 / sum(pca_base$sdev^2),
  cum_variance = cumsum(pca_base$sdev^2 / sum(pca_base$sdev^2))
)

knitr::kable(var_explained, digits = 3,
             caption = "Variance Explained (manual)")
```

The manual biplot requires extracting matrices, scaling loadings to match score
ranges, computing variance percentages for axis labels, and assembling the
ggplot layer by layer. `tidy_pca_biplot()` handles all of that in one call.
And the `kable()` variance table is functional but plain — no colour coding, no
cumulative-variance highlighting. `tl_table_variance()` adds these by default.

---

## 2. Hierarchical Clustering

Hierarchical clustering involves computing distances, fitting the tree,
visualising the dendrogram, cutting it, and augmenting your data with cluster
assignments. Each step traditionally produces a different data structure.

### With tidylearn

```{r tl-hclust}
hc <- tidy_hclust(USArrests, method = "ward.D2")

# Dendrogram with cluster rectangles
tidy_dendrogram(hc, k = 4)
```

Cut the tree and get a tidy tibble of cluster assignments:

```{r tl-hclust-cut}
clusters <- tidy_cutree(hc, k = 4)
knitr::kable(head(clusters, 10), caption = "Cluster Assignments (first 10)")
```

Augment the original data and produce a formatted cluster summary table:

```{r tl-hclust-table}
hc_model <- tl_model(USArrests, method = "hclust")
tl_table_clusters(hc_model, k = 4)
```

### Without tidylearn

```{r base-hclust}
# Compute distance matrix
d <- dist(scale(USArrests), method = "euclidean")

# Fit hierarchical clustering
hc_base <- hclust(d, method = "ward.D2")

# Plot dendrogram
plot(hc_base, main = "Hierarchical Clustering Dendrogram",
     xlab = "", sub = "", cex = 0.7)
rect.hclust(hc_base, k = 4, border = 2:5)
```

The dendrogram looks similar — both use base R graphics for this. The real
difference is in what happens next:

```{r base-hclust-cut}
# cutree returns a named integer vector — not a tibble
clusters_base <- cutree(hc_base, k = 4)
str(clusters_base)
```

```{r base-hclust-augment}
# To get a summary table, manually bind and reshape
USArrests_clustered <- USArrests
USArrests_clustered$cluster <- clusters_base

USArrests_clustered %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), mean), .groups = "drop") %>%
  knitr::kable(digits = 1, caption = "Cluster Means (manual)")
```

The dendrogram itself is comparable. But `cutree()` returns a named integer
vector that needs manual binding to your data, and the resulting `kable()` is
plain text. `tl_table_clusters()` produces a formatted table with cluster sizes,
styled headers, and consistent theming — ready for a report.

---

## 3. Lasso Regularisation: Coefficient Paths and Tables

Regularised models are a common choice, but visualising how coefficients shrink
along the regularisation path is one of those tasks where the default output is
decidedly not report-ready.

### With tidylearn

```{r tl-lasso}
lasso <- tl_model(mtcars, mpg ~ ., method = "lasso")

# Coefficient path as a ggplot2 object
tl_plot_regularization_path(lasso)
```

```{r tl-lasso-cv}
# Cross-validation curve
tl_plot_regularization_cv(lasso)
```

And a formatted coefficient table, sorted by magnitude with zero coefficients
greyed out:

```{r tl-lasso-table}
tl_table_coefficients(lasso)
```

Both plots are ggplot2 objects — `theme_minimal()`, consistent aesthetics, and
directly passable to `ggplotly()` or `ggsave()`. The table is a `gt` object
with the same consistent styling.

### Without tidylearn

```{r base-lasso}
library(glmnet)

# Prepare model matrix (glmnet doesn't accept formulas)
x <- model.matrix(mpg ~ ., data = mtcars)[, -1]
y <- mtcars$mpg

# Fit with cross-validation
cv_fit <- cv.glmnet(x, y, alpha = 1)
```

The default coefficient path plot:

```{r base-lasso-path}
plot(cv_fit$glmnet.fit, xvar = "lambda", label = TRUE)
```

The default cross-validation plot:

```{r base-lasso-cv}
plot(cv_fit)
```

These are base R graphics — functional, but they can't be themed, faceted,
combined with other ggplot2 panels, or converted to interactive plotly charts.
Building ggplot2 equivalents from the `glmnet` object requires extracting the
coefficient matrix across all lambda values and pivoting it to long format:

```{r base-lasso-gg}
# Extract coefficient matrix
coef_matrix <- as.matrix(cv_fit$glmnet.fit$beta)
lambda_vals <- cv_fit$glmnet.fit$lambda

# Reshape to long format for ggplot
coef_df <- as.data.frame(t(coef_matrix))
coef_df$lambda <- lambda_vals
coef_long <- tidyr::pivot_longer(coef_df, -lambda,
                                  names_to = "variable",
                                  values_to = "coefficient")

ggplot(coef_long, aes(x = log(lambda), y = coefficient, colour = variable)) +
  geom_line() +
  labs(title = "Lasso Coefficient Path", x = "log(lambda)",
       y = "Coefficient", colour = "Variable") +
  theme_minimal()
```

```{r base-lasso-coef-table}
# Extract coefficients at lambda.1se — returns a sparse matrix
coefs <- as.matrix(coef(cv_fit, s = "lambda.1se"))
coef_tbl <- data.frame(
  term = rownames(coefs),
  estimate = as.vector(coefs)
)
coef_tbl <- coef_tbl[order(-abs(coef_tbl$estimate)), ]

knitr::kable(coef_tbl, digits = 4, row.names = FALSE,
             caption = "Lasso Coefficients at lambda.1se (manual)")
```

The manual approach works, but it's the kind of reshaping code that's easy to
get subtly wrong and tedious to repeat. `tl_plot_regularization_path()` handles
extraction, pivoting, labelling, and theming in one call. And the `kable()`
coefficient table is plain — `tl_table_coefficients()` adds sorting, zero
greying, and the selected lambda value in the subtitle.

---

## 4. Multi-Model Classification Comparison

Comparing models across packages is where consistent output structure matters
most. Each package has its own prediction interface, metric accessors, and
plot conventions.

### With tidylearn

```{r tl-comparison}
split <- tl_split(iris, prop = 0.7, stratify = "Species", seed = 42)

# Fit three models — same interface for each
m_logistic <- tl_model(split$train, Species ~ ., method = "logistic")
m_forest   <- tl_model(split$train, Species ~ ., method = "forest")
m_tree     <- tl_model(split$train, Species ~ ., method = "tree")
```

A formatted comparison table — one call:

```{r tl-comparison-table}
tl_table_comparison(
  m_logistic, m_forest, m_tree,
  new_data = split$test,
  names = c("Logistic", "Random Forest", "Decision Tree")
)
```

And a confusion matrix for any model:

```{r tl-confusion-table}
tl_table_confusion(m_forest, new_data = split$test)
```

Adding a fourth model is just another argument in `tl_table_comparison()` — the
table code stays unchanged.

### Without tidylearn

```{r base-comparison}
library(randomForest)

set.seed(42)
train_idx <- unlist(lapply(
  split(seq_len(nrow(iris)), iris$Species),
  function(i) sample(i, size = floor(0.7 * length(i)))
))
train_data <- iris[train_idx, ]
test_data  <- iris[-train_idx, ]

# Three different packages, three different interfaces
fit_glm  <- glm(Species ~ ., data = train_data, family = "binomial")
fit_rf   <- randomForest(Species ~ ., data = train_data)
fit_tree <- rpart::rpart(Species ~ ., data = train_data, method = "class")

# Each returns predictions in a different format
pred_glm  <- predict(fit_glm, newdata = test_data, type = "response")
pred_rf   <- predict(fit_rf, newdata = test_data)
pred_tree <- predict(fit_tree, newdata = test_data, type = "class")

# glm doesn't handle multiclass — can't even compute accuracy
acc_glm  <- NA_real_
acc_rf   <- mean(pred_rf == test_data$Species)
acc_tree <- mean(pred_tree == test_data$Species)

comparison_base <- data.frame(
  model = c("Logistic (GLM)", "Random Forest", "Decision Tree"),
  accuracy = c(acc_glm, acc_rf, acc_tree)
)

knitr::kable(comparison_base, digits = 3,
             caption = "Model Comparison (manual)")
```

Three different packages, three different prediction interfaces, and `glm()`
can't even handle the multiclass case without implementing one-vs-rest manually.
The `kable()` output is plain and limited to one metric. With tidylearn,
`tl_table_comparison()` produces a styled, multi-metric comparison table and
`method = "logistic"` handles multiclass dispatching internally.

---

## 5. Interactive Reporting with plotly

Because tidylearn's plot functions return standard ggplot2 objects, converting
any visualisation to an interactive plotly chart is a one-liner:

```{r plotly-example, eval = FALSE}
library(plotly)

# tidylearn's plot returns a ggplot2 object — pass it straight to ggplotly
ggplotly(tidy_pca_biplot(pca, label_obs = TRUE))
ggplotly(tl_plot_regularization_path(lasso))
ggplotly(plot(m_forest, type = "confusion"))
```

`ggplotly()` picks up axis labels, themes, and tooltip data automatically.
Compare this to the base R plots from `biplot()`, `plot.glmnet()`, or
`plot.hclust()` — none of which can be converted to plotly without rebuilding
them from scratch.

---

## Why This Matters

**Consistent, polished output by default.** Every model — whether it's a PCA
biplot, a lasso coefficient path, or a confusion matrix — returns ggplot2
plots and `gt` tables with a consistent visual language. You don't need to
learn each package's idiosyncratic output format or build custom formatting
code to get report-quality visuals and tables.

**Reproducibility through uniformity.** When your reporting pipeline works the
same way for every model type, your analysis becomes genuinely reproducible.
Swap `method = "forest"` for `method = "xgboost"` and rerun — the same
`tl_table()` calls, the same `plot()` calls, the same comparison logic all
work without modification. That means you can iterate on model selection
without touching your reporting layer, and anyone reading your code can follow
the same pattern across different analyses.

The best analysis code is code that gets out of your way and lets you focus on
the results. That's what tidylearn is for.
